---
title: 'Assignment 1'
author: 'Andrii Voitkiv'
date: "January 22th, 2023"
output:
  github_document:
    math_method: webtex
    html_preview: false
    toc: true
    toc_depth: 2
---

```{r global_options}
knitr::opts_chunk$set(fig.path='Figs/')
```
## Question 1
```{r}
# Read the dataset
data <- read.csv("/Users/berg/DataspellProjects/MDSA-UofC/DATA606/01_assignment/ks-projects-201801.csv", header = TRUE, sep = ",")
```

```{r}
# Check the dimensions of the dataset and column names
dim(data)
colnames(data)
```


## Question 2
```{r}
# Simple random sample of 5000 observations
set.seed(2023)
ix <- sample(1:nrow(data), 5000, replace = FALSE)
srs <- data[ix,]
# Calculate the sample mean and its standard deviation
srs_mean = mean(srs$usd_pledged_real)
srs_sd = sd(srs$usd_pledged_real)
# Print the results
print(paste("The sample mean is", srs_mean))
print(paste("The sample standard deviation is", srs_sd))
```
```{r}
# Method 1: Estimate the population mean and its standard deviation using survey package
library("survey")
# Simple random sample of 5000 observations
set.seed(2023)
N = nrow(data)
n = 5000
ix <- sample(1:N, n, replace = FALSE)
srs <- data[ix,]
# Add a sampling weight column and finite population correction factor to the sample
srs = data.frame(srs, pw = rep(N/n, n), fpc = rep(N, n))
svy1 = svydesign(id = ~0, strata = NULL, data = srs, weights = ~pw, fpc = ~fpc)
# Estimate the population mean and its standard deviation
srs_mean = svymean(~usd_pledged_real, design = svy1)
# Print the results
srs_mean
```

```{r}
# Calculate confidence interval for the population mean of usd_pledged_real
confint(srs_mean ,level = 0.95, df=degf(svy1))
```
Conclusion: From the sample taken without replacement the svydesign function in the survey package estimates the population mean to be 8227 USD and the . We can be 95% confident that the true population mean is in the range between 6659 USD and 9797 USD.

```{r}
# Method 2: Estimate the population mean and its standard deviation using simulation
mean_SD<-function(s, n){  # s denotes how many times you draw the sample, n denotes the sample size
  set.seed(2023)
  sam_mean<-rep(0, s)
  sam_sd <- rep(0, s)
  for (i in 1:s){
    ind = sample(seq(0, N), n, replace=FALSE)
    sam_mean[i] = mean(data$usd_pledged_real[ind])
    sam_sd[i] = sd(data$usd_pledged_real[ind])
  }
  es_mean <- mean(sam_mean)
  es_var <- sam_sd[1]^2/n*(1-n/N)
  es_var2 <- sd(sam_mean)^2
  print(c(es_mean,es_var,es_var2))
  # Plot the histogram of the sample mean using ggplot2
  library(ggplot2)
ggplot(data.frame(sam_mean), aes(x=sam_mean)) +
    geom_histogram(binwidth=500) +
    geom_vline(xintercept=mean(data$usd_pledged_real), color="red", size=1) +
    geom_text(x=mean(data$usd_pledged_real), y=0, label="Population mean", vjust=-1) +
    labs(title="Histogram of the sample mean", x="Sample mean", y="Frequency")
}
# Run the function
mean_SD(50, 5000)

```
Conclusion: The method of simulating simple random sample without replacement estimates the population mean to be 9198 USD and population variance 689451.448 (with correction factor) or 1527613 (without correction factor). The histogram of the sample mean is plotted above. The red line indicates the population mean.

## Question 3
```{r}
# Calculate proportion of each category and make a data frame
prop = table(data$main_category)/nrow(data)
# Sort the data in data frame order
prop = prop[unique(data$main_category)]
# Create a vector of the proportion of each category
prop = as.vector(prop)
# Calculate the sample size for each strata
strata_size = round(prop * 5000)
# Test if the sum of the sample size equals 5000
print(paste("The sample size is:", sum(strata_size)))
# Print the strata sizes
print(strata_size)
```
```{r}
library("sampling")
set.seed(2023)
# Sampling strata_size observations from each stratum with simple random sampling
idx = sampling::strata(data, stratanames = "main_category", size = strata_size, method = "srswor")
# Create a new data frame with the sampled observations
strata_srs = getdata(data, idx)
# Add columns for sampling weight and finite population correction factor
strata_srs = data.frame(strata_srs, pw = rep(N/nrow(strata_srs), nrow(strata_srs)), fpc = c(rep(39874,527),rep(63585,840),rep(51918,686), rep(24602, 325), rep(30070, 397), rep(8809, 116), rep(35231, 465), rep(10819, 143), rep(22816, 301), rep(10913, 144), rep(28153, 372), rep(10779, 142), rep(32569, 430), rep(3768, 50), rep(4755, 63)))
# Estimate the population mean and its standard deviation
svy2 = svydesign(id = ~1, strata = ~main_category, data = strata_srs, weights = ~pw, fpc = ~fpc)
me2 = svymean(~usd_pledged_real, svy2)
# Print the results
print(me2)
```

```{r}
# Calculate confidence interval for the population mean of usd_pledged_real
confint(me2, level = 0.95, df=degf(svy2))
```
Conclusion: From the stratified sample the svydesign function in the survey package estimates the population mean to be 9889 USD. We can be 95% confident that the true population mean is in the range between 6867 USD and 12911 USD.

## Question 4
```{r}
# Calculate the number of unique clusters in the data
n_clusters = length(unique(data$country))
# Print unique values of the country column
print(unique(data$country))
# Print the length of the unique values
print(paste("Number of clusters (countries) in the dataset:", n_clusters))
```

```{r}
library("sampling")
library("survey")
# Select two clusters with simple random sampling method using the sampling package
set.seed(2023)
n_choose = 2
idx = sampling::cluster(data, clustername = "country", size = n_choose, method = "srswor")
# Create a new data frame with the sampled observations
cluster_srs = getdata(data, idx)
# Print the clusters (countries) in the sample
print(paste(
    "The clusters (countries) in the sample are:",
    paste(unique(cluster_srs$country), collapse = ", ")
    ))
```
```{r}
# Calculate the sampling weight and finite population correction factor
cluster_srs = data.frame(cluster_srs, pw = rep(n_clusters/n_choose, nrow(cluster_srs)), fpc = rep(n_clusters, nrow(cluster_srs)))
# Estimate the population mean and its standard deviation
svy3 = svydesign(id = ~country, data = cluster_srs, weights = ~pw, fpc = ~fpc)
me3 = svymean(~usd_pledged_real, svy3)
# Print the results
print(me3)
```

```{r}
# Calculate confidence interval for the population mean of usd_pledged_real
confint(me3, level = 0.95, df=degf(svy3))
```
Conclusion: From the cluster sample the svydesign function in the survey package estimates the population mean to be 4516 USD and the standard deviation 2985.4. We can be 95% confident that the true population mean is in the range between 0 USD and 42,448 USD.

## Question 5
### 1. Use the simple random sample you get from Q2 to estimate the ratio between the actual amount of funding pledged and the goal amount of funding and its standard deviation
```{r}
# Calculate the ratio between the actual amount of funding pledged and the goal amount of funding
ratio = svyratio(~usd_pledged_real, ~usd_goal_real, svy1)
# Print the results
print(ratio)
```

### 2. Estimate how much you can get on average based on your goal amounts and give the standard deviation of your estimate
```{r}
# Make a vector of goal amounts
goal_amount = c(500000, 1000000, 2000000)
# Predict the average amount of funding pledged based on the goal amount of funding
pred1 = predict(ratio, total = sum(goal_amount))
# Print the results
print(paste("The average amount of funding pledged based on the goal amount of funding of 500000 USD is:", pred1$total / length(goal_amount),
            "with standard deviation", pred1$se / length(goal_amount)))
```

