---
title: "Rescue Time"
subtitle: "An app to track your computer and mobile phone usage"
author:
    - Andrii Voitkiv
date: "`r format(Sys.time(), '%a, %b %d, %Y')`"
geometry: margin=2cm
output:
    html_document:
        toc: true
        toc_float: false
        toc_depth: 3
        number_sections: false
        fig_width: 12
        fig_height: 8
        fig_retina: 6
        fig_path: Figs/
        keep_md: false
        theme: united
        highlight: tango # pygments, kate, espresso, zenburn, haddock, textmate

---


```{r global_options, echo=FALSE}
# Set global options
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results='asis')
```

_Creating your Story:_

_Structure your data story around a set of exactly 5 clear and explicit analysis questions._
_For each analysis question, describe the analysis process you used to try to answer it._
_Provide a concise discussion of your findings, supported by clear charts, statistics, and other analytic results. Include clear, well-captioned images that illustrate your analysis process, findings, and any problems you encountered. Annotating the images with captions, arrows, and notes can be very helpful._

```{r read_data, cache=TRUE}
# Read data
RT_mobile <- read.csv("/Users/berg/DataspellProjects/MDSA-UofC/DATA605/10_project/data/rescuetime-minute-grouped-mobile.csv")
RT_pc <- read.csv("/Users/berg/DataspellProjects/MDSA-UofC/DATA605/10_project/data/rescuetime-minute-grouped-pc.csv")
```

```{r}
# Productivity colors from Rescue Time app:----
Productivity_col <- rgb(red = c(13, 65, 177, 218, 212, 160),
                        green = c(88, 130, 193, 105, 28, 235),
                        blue = c(193, 221, 191, 93, 21, 142),
                        names = c("Very Productive", "Productive", "Neutral",
                                  "Distracting", "Very Distracting", "Mobile Phone"),
                        maxColorValue = 255)
colors <- Productivity_col[c("Productive", "Very Distracting")]
```

```{r}
# Use the ggridges package to create a density plot. On x-axis we have the 'hm' variable, which is the hour and minute of the day. On the y-axis use day names. Plot both mobile and pc Duration.
RT_mobile$weekday <- factor(RT_mobile$weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
RT_pc$weekday <- factor(RT_pc$weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
```

```{r}
# Combine the two data frames into one data frame with a Device column that I lost during the grouping process
RT_mobile$Device <- "Mobile"
RT_pc$Device <- "PC"
RT_mobile_pc <- rbind(RT_mobile, RT_pc)
# Make `Device` a factor
RT_mobile_pc$Device <- factor(RT_mobile_pc$Device, levels = c("Mobile", "PC"))
```

```{r}
library(dplyr)
# Group by Device, weekday and hm. Summarize Duration with mean
RT_mobile_pc <- RT_mobile_pc %>%
  group_by(Device, weekday, hm) %>%
  summarize(Duration = mean(Duration)) %>%
  mutate(row = row_number() - 1)
```
```{r}
# Create a vector of breaks for the x-axis
starttime <- 6
time_interval <- 5
breaks <- seq(0, (24 - starttime) * (60 / time_interval) - 1, by = 60 / time_interval)

# Filter data by `row` more than 72 (6 am) = 6 * 12 = 72
RT_mobile_pc <- RT_mobile_pc %>%
  filter(row > starttime * (60 / time_interval) - 1)

# Update row variable to start from 0
RT_mobile_pc$row <- RT_mobile_pc$row - starttime * (60 / time_interval)
```




```{r}
# Make a ggridges plot using the ggridges package
library(ggridges)
library(ggplot2)
library(ggtext)

RT_mobile_pc_averages <- RT_mobile_pc %>%
  group_by(weekday, Device) %>%
  summarize(average = round(mean(Duration))) %>%
  mutate(average_hours = floor(average / 60),
         average_minutes = average %% 60,
         average_formatted = paste0(average_hours, "h ", average_minutes))


# Create a title and subtitle
title <- paste0("Time spent on <span style = 'color: ",
                colors[1], ";'>", levels(RT_mobile_pc$Device)[1],
                "</span> vs. <span style = 'color: ",
                colors[2], ";'>", levels(RT_mobile_pc$Device)[2], "</span>")
subtitle <- paste0("Meausred in minutes by Rescue Time app")

# make the plot:
ggplot(data = RT_mobile_pc, aes(x = row, y = weekday)) +
  geom_density_ridges(aes(fill = Device, color = Device,  height = Duration),
                      stat="identity", alpha = 0.75, size = 0.2, scale = 0.9, show.legend = FALSE) +
  geom_text(data = RT_mobile_pc_averages, aes(x= (24 - starttime) * (12 + 1/4), y = c(1:7 + 0.5, 1:7 + 0.2),
                               label = paste(average_formatted,  "/day"),
                               hjust = 0, vjust = 0, family = 'Comfortaa', color = Device),
            show.legend = FALSE, size = 3.5) +
  labs(x = "Time of Day", y = NULL, title = title, subtitle = subtitle) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_family = "Comfortaa") +
  theme(plot.margin = unit(c(2, 22, 2, 2), "mm"),
        plot.background = element_rect(fill='transparent', color=NA),
        panel.grid = element_blank(),
        plot.title.position = "plot",
        plot.title = element_markdown(face = "bold", size = 15, hjust = 0, color = "grey10"),
        plot.subtitle = element_markdown(size = 11, hjust = 0, color = "grey20", margin = margin(b = 20)),
        axis.title.x = element_text(face = "bold", margin = margin(t = 6), color = "grey40", size = 10),
        axis.text.x = element_text(size = 8, color = "grey50", margin = margin(t = 5)),
        axis.text.y = element_text(size = 10, color = "grey30", margin = margin(r = 5), vjust = 0),
        axis.ticks.x = element_line(color = "grey60", linewidth = 0.2)) +
  scale_x_continuous(expand = c(0, 0), breaks = breaks, labels = RT_mobile_pc$hm[breaks + 1]) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_manual(values = hsv(h = rgb2hsv(col2rgb(colors))[1,],
                                 s = rgb2hsv(col2rgb(colors))[2,],
                                 v = rgb2hsv(col2rgb(colors))[3,])) +
  scale_color_manual(values = hsv(h = rgb2hsv(col2rgb(colors))[1,],
                                  s = rgb2hsv(col2rgb(colors))[2,],
                                  v = rgb2hsv(col2rgb(colors))[3,] - 0.1))

```





