---
title: "Rescue Time"
subtitle: "An app to track your computer and mobile phone usage"
author:
    - Andrii Voitkiv
date: "`r format(Sys.time(), '%a, %b %d, %Y')`"
geometry: margin=3cm
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 3
        number_sections: false
        fig_width: 12
        fig_height: 8
        fig_retina: 6
        fig_path: Figs/
        keep_md: false
        theme: united
        highlight: tango # pygments, kate, espresso, zenburn, haddock, textmate

---


```{r global_options, echo=FALSE}
# Set global options
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results='asis')
```
```{r}
library('dplyr')
library('tidyverse')
```
_Creating your Story:_

_Structure your data story around a set of exactly 5 clear and explicit analysis questions._
_For each analysis question, describe the analysis process you used to try to answer it._
_Provide a concise discussion of your findings, supported by clear charts, statistics, and other analytic results. Include clear, well-captioned images that illustrate your analysis process, findings, and any problems you encountered. Annotating the images with captions, arrows, and notes can be very helpful._

```{r read_data, cache=TRUE}
# Read data
RT_mobile <- read.csv("/Users/berg/DataspellProjects/MDSA-UofC/DATA605/10_project/data/rescuetime-minute-grouped-mobile.csv")
RT_pc <- read.csv("/Users/berg/DataspellProjects/MDSA-UofC/DATA605/10_project/data/rescuetime-minute-grouped-pc.csv")
```
```{r}
# Productivity colors from Rescue Time app:----
Productivity_col <- rgb(red = c(13, 65, 177, 218, 212, 160),
                        green = c(88, 130, 193, 105, 28, 235),
                        blue = c(193, 221, 191, 93, 21, 142),
                        names = c("Very Productive", "Productive", "Neutral",
                                  "Distracting", "Very Distracting", "Mobile Phone"),
                        maxColorValue = 255)
colors <- Productivity_col[c("Productive", "Very Distracting")]
```
```{r}
# Make a ggridges plot using the ggridges package
library(ggridges)
library(ggplot2)
# Use the ggridges package to create a density plot. On x-axis we have the 'hm' variable, which is the hour and minute of the day. On the y-axis use day names. Plot both mobile and pc Duration.
RT_mobile$weekday <- factor(RT_mobile$weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
RT_pc$weekday <- factor(RT_pc$weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
```


```{r}
# Group by
```

```{r}
# Combine the two data frames into one data frame with a Device column that I lost during the grouping process
RT_mobile$Device <- "Mobile"
RT_pc$Device <- "PC"
RT_mobile_pc <- rbind(RT_mobile, RT_pc)

```
```{r}
# Group by Device, weekday and hm. Summarize Duration with mean
RT_mobile_pc <- RT_mobile_pc %>%
  group_by(Device, weekday, hm) %>%
  summarize(Duration = mean(Duration)) %>%
  mutate(row = row_number() - 1)
```



```{r}
ggplot(data = RT_mobile_pc, aes(x = row, y = weekday)) +
  geom_density_ridges(aes(fill = Device, color = Device,  height = Duration),
                      stat="identity", alpha = 0.75, size = 0.2, scale = 0.9, show.legend = FALSE) +
  coord_cartesian(clip = "off") +
  labs(x = "Time of Day", y = NULL, title = "title", subtitle = "subtitle") +
  theme_minimal(base_family = "Comfortaa")

```


```{r}
# Column names RT_mobile
colnames(RT_mobile)
```

```{r}
```

